---
title: "US Weather Effects on Population Health and the Economic Consequences"
output: html_document
---

## Synopsis
In this report we aim to describe which type of US weather events are most
harmful with respect to population health (injuries and fatalities) and which
have the greatest economic consequences (property and crop damage).
Exploring the NOAA Storm Database, containing data from 1950 through
November 2011, we observe that XXXX  XXXXX.

--------------------------------------------------------------------------------

## Data Processing
The following R libraries are used.
```{r}
library(stringr)
library(dplyr)
library(ggplot2)
```

### Reading in the data
We first read in the data from the csv file in
[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
47mb bzip2 file.
The data is a bzip2 compressed, comma separated file containing a header and
missing values are coded as blank fields.
```{r noaa-severe-weather-read, cache=TRUE}
storms.orig <- read.csv("data/repdata-data-StormData.csv.bz2", header=TRUE,
                        na.strings="", stringsAsFactors=FALSE)
```

After reading the data into `storms.orig` we see that there are
`r nrow(storms.orig)` observations (rows) and 
`r ncol(storms.orig)` variables (columns) in this dataset.
```{r}
dim(storms.orig)
```

#### Exploring the data
We explore the first few rows in this dataset.  These documents help us find how some of the variables are constructed/defined.

  - National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
  - National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r}
head(storms.orig)
```

The `EVTYPE` event type variable will be needed to determine the weather event.  Besides EVTYPE, the following variables contain data useful for this report.

  - To determine population health, the following variables are needed:
    - `FATALITIES` - The number of direct or indirect fatalities attributable to the event.
    - `INJURIES` - The number of direct or indirect injuries attributable to the event.
  - Economic consequences:
    - `PROPDMG` - Property damage dollar amount
    - `PROPDMGEXP` - Property damage dollar amount magnitude (K, M, B, etc.)
    - `CROPDMG` - Crop damage dollar amount
    - `CROPDMGEXP` - Crop damage dollar amount magnitude (K, M, B, etc.)

#### Subsetting the dataset
Reducing the dataset to only those variables we are focusing on both helps
pre-eliminate non-essential values that we would have to otherwise visually
inspect later in the analysis and also improves subsequent computational performance.

`storms.df` will contain the reduced, cleaned, and tidied dataset used
throughout the remainder of the report.
```{r}
storms.df <- storms.orig[,c("EVTYPE",
                            "FATALITIES", "INJURIES",
                            "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
dim(storms.df)
```
The number of variables used decreases from `r ncol(storms.orig)`
to `r ncol(storms.df)`.

#### Filtering the dataset
Reducing the dataset to only those observations where there are injures/fatalities
or property/crop damage also helps.
```{r}
storms.df <- storms.df[storms.df$FATALITIES>0 | storms.df$INJURIES>0 |
                       storms.df$PROPDMG>0 | storms.df$CROPDMG>0, ]
dim(storms.df)
```
The number of observations used decreases from `r nrow(storms.orig)`
to `r nrow(storms.df)`.

#### Tidying Event Types
The previous subsetting and filter results in `r length(unique(storms.df$EVTYPE))`
unique `EVTYPE` values we need to consider in the reduced `storms.df` dataset
versus the original `r length(unique(storms.orig$EVTYPE))` values in `storms.orig`.
```{r}
unique(storms.df$EVTYPE)
```

Still, there are many event types.  Through visual inspection, we'll now attempt
to somewhat normalize them while retaining a reasonable level of fidelity.
The code below shows the series of transformations used.
```{r}
normalizedEVTYPE <- storms.df$EVTYPE

# Upper case all
normalizedEVTYPE <- toupper(normalizedEVTYPE)

# Replace special characters /, (, ), ., \, &, - with space
normalizedEVTYPE <- gsub("/|\\(|\\)|\\.|\\\\|\\&|\\-", " ", normalizedEVTYPE)

# Remove numeric F, G, or numeric magnitudes
normalizedEVTYPE <- gsub("[F|G]\\d+", " ", normalizedEVTYPE)
normalizedEVTYPE <- gsub("\\d+", " ", normalizedEVTYPE)

# Remove plurality
normalizedEVTYPE <- gsub("RAINS\\s", "RAIN ", normalizedEVTYPE)
normalizedEVTYPE <- gsub("STORMS", "STORM", normalizedEVTYPE)
normalizedEVTYPE <- gsub("WINDS", "WIND", normalizedEVTYPE)
normalizedEVTYPE <- gsub("FLOODS", "FLOOD", normalizedEVTYPE)

# Remove adjectives
normalizedEVTYPE <- gsub("HARD |HIGH |SEVERE |HEAVY |EXCESSIVE |STRONG | INJURY", " ", normalizedEVTYPE)

# Any tornado
normalizedEVTYPE <- gsub("TORNADOES|TORNDAO", "TORNADO", normalizedEVTYPE)
normalizedEVTYPE <- gsub(".*TORNADO.*", "TORNADO", normalizedEVTYPE)

# Any hurricane
normalizedEVTYPE <- gsub(".*HURRICANE.*", "HURRICANE", normalizedEVTYPE)

# Any thunderstorm
normalizedEVTYPE <- gsub("TSTM|THUNERSTORM|TUNDERSTORM|THUDERSTORM|THUNDERTORM|THUNDEERSTORM", "THUNDERSTORM", normalizedEVTYPE)
normalizedEVTYPE <- gsub(".*THUNDERSTORM.*", "THUNDERSTORM", normalizedEVTYPE)

# Snow
normalizedEVTYPE <- gsub("SNOWFALL", "SNOW", normalizedEVTYPE)

# Trim adjacent whitespace
normalizedEVTYPE <- gsub("\\s+", " ", normalizedEVTYPE)

# Trim leading and trailing blanks
normalizedEVTYPE <- str_trim(normalizedEVTYPE)

storms.df$normalizedEVTYPE <- normalizedEVTYPE
unique(normalizedEVTYPE)
length(unique(normalizedEVTYPE))
```

The number of unique types has been reduced from the
original `r length(unique(storms.orig$EVTYPE))`
to `r length(unique(normalizedEVTYPE))`.
```{r}
unique(normalizedEVTYPE)
```

--------------------------------------------------------------------------------

## Results

### Injuries and Fatalities
Copy and filter the reduced `storms.df` dataset to only those observations where
there is an injury or fatality and store it in `human.df`.
Summarize the injuries and fatalities by event type and store it in `human.summary`.
```{r}
human.df <- storms.df[storms.df$FATALITIES>0 | storms.df$INJURIES>0, ]
human.summary <- human.df %>% group_by(normalizedEVTYPE) %>%
    summarise(souls=sum(INJURIES, FATALITIES))
human.total <- sum(human.summary$souls)
```
There are a total of *`r as.character(human.total)`* injuries and fatalities.

Plot `human.summary` and notice the event with the worst impact.
```{r}
# Find event with worst impact
index <- which.max(human.summary$souls)
evtype <- human.summary$normalizedEVTYPE[index]
cnt <- human.summary$souls[index]

# Plot
ggplot(human.summary, aes(x=normalizedEVTYPE, y=souls)) +
  ggtitle("Injuries and Fatalities per Event Type") +
  xlab("Event Type") + ylab("Count") +
  theme_bw() +
  geom_point() +
  geom_text(hjust=1, yjust=0,
            aes(label=paste0(evtype, ", ", cnt),
                x=human.summary$normalizedEVTYPE[index-2], y=cnt))
```

### Property and Crop Damage
```{r}
damage.df <- storms.df[storms.df$PROPDMG>0 | storms.df$CROPDMG>0, ]
dim(damage.df)
```